---
title: "Lab2- git, GitHub and Rstudio projects"
author: "Briana Kubik"
date: "9/13/2020"
output: html_document
bibliography: data/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercises for Lab 1
Document and save R script file (Lab1_BrianaKubik.R).

### Exercise 1
For x = 3 and y = 15, compute the sum, difference, product, and quotient of x and y.

```{r ex1}
x = 3
y = 15

#Sum
sum(x,y)

#Difference
x - y

#Product
x * y

#Quotient
x / y
```

### Exercise 2
For x = 2 and y = 3, test x^5 is greater than y^4 and return a logical value.

```{r ex2}
x = 2
y = 3
x^5 > y^4
```

### Exercise 3
Create a vector of the values 22, 62, 148, 43, and 129. Determine the sum of the vector. Divide each value in the vector by the sum to determine relate frequency.

```{r ex3}
vec = c(22, 62, 148, 43, 129)
sum_vec = sum(vec)
vec / sum_vec
```

### Exercise 4 
Create a vector of the nucleotides A, T, C, and G. Sort vector.

```{r}
vec2 = c("A", "T", "C", "G")
sort(vec2)
```

Application written in R [@RCoreTeam] using the Shiny framework [@Chang2015]. 

# References