---
title: "Lab 3A- Starting with Data"
author: "Briana Kubik"
date: "9/7/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning = FALSE}
#install.packages("tidyverse")
library(tidyverse)
```

# Presentation of the survey data

The R function `download.file()` is used to download a file from the internet onto your local machine (in this case, it is a .csv file from Figshare). The argument specified by `destfile = ` is the path to where the file will be saved on your local computer.
```{r eval = FALSE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

Once we have downloaded the data, we can assign it a variable and view it in R. The function `read.csv` assumes that the data are delineated by commas. In some formats, a semicolon (;) is used as the delineator so the function `read.csv2()` can be used. The `read.delim()` function can be used for data that are tab delineated. All of these are versions of the `read.table()` function which can specify specific delineators (i.e. read.table(file = "data/portal_data_joined.csv", sep = ",", header = TRUE)).
```{r}
surveys = read.csv("data/portal_data_joined.csv") #read in data and assign to a variable called "surveys"
head(surveys) #shows only the 1st 6 rows of the data
#View(surveys) #opens full data file in new R page
```

# What are data frames

Data frames are a common way to present tabular data. The columns are vectors that all must have the same length and type of data. This information can be found using a function called `str()`:
```{r}
str(surveys) #shows the structure of the data file
```

# Inspecting `data.frame` objects

These are some others ways to gather information about a dataset:
```{r results = 'hide'}
dim(surveys) #dimensions of the data frame
nrow(surveys) #number of rows in the data frame
ncol(surveys) #number of columns in the data frame
head(surveys) #shows first 6 rows of data frame
tail(surveys) #shows last 6 rows of data frame
names(surveys) #shows column names
rownames(surveys) #shows row names
str(surveys) #returns the structure of the object
summary(surveys) #returns summary statistics on each column
```


### Challenge 1

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?\
**Data frame**
* How many rows and how many columns are in this object?\
**34786 rows and 13 columns**
* How many species have been recorded during these surveys?\
**48 species**

```{r}
str(surveys)
```

# Indexing and subsetting data frames

Data frames are organized like matrices where each value has a set of coordinates so they can be individually specified. The general syntax for specifying coordinates in R is [*row*, *column*].

```{r results = 'hide'}
surveys[1, 1] #first element in the first column of the data frame (as a vector)
surveys[1, 6] #first element in the 6th column (as a vector)
surveys[, 1] #first column of the data frame (as a vector)
surveys[1] #first column of the data frame (as a data.frame)
surveys[1:3, 7] #first three elements in the 7th column (as a vector)
surveys[3, ] #the 3rd row of the data frame (as a data.frame)
head_surveys = surveys[1:6, ] #equivalent to head_surveys = head(surveys)
```

The `:` operator will create a range of integers in ascending or descending order. You can also exclude numbers using `-`.

```{r results = 'hide'}
surveys[, -1] #the whole data frame, except the first column
surveys[-(7:34786), ] #equivalent to head(surveys)
```

You can also subset data frames based on their column names

```{r results = 'hide'}
surveys["species_id"] #result is a data.frame
surveys[, "species_id"] #result is a vector
surveys[["species_id"]] #result is a vector
surveys$species_id #result is a vector
```

### Challenge 2

1. Create a `data.frame` (`survey_200`) containing only the data in row 200 of the `surveys` dataset. 

```{r}
surveys_200 = surveys[200,]
```

2. Notice how `nrow` gave you the number of rows in a `data.frame`?

* Use that number to pull out the last row in the data frame.
* Compare that with what you see as the last row using `tail()` to make sure it’s meeting expectations.
* Pull out that last row using `nrow()` instead of the row number.
* Create a new data frame (`surveys_last`) from that last row.

```{r}
nrow(surveys)
surveys[34786,]
tail(surveys)
surveys[nrow(surveys),]
surveys_last = surveys[nrow(surveys),]
```

3. Use `nrow()` to extract the row that is in the middle of the data frame. Store the content of this row in an object named `surveys_middle`.

```{r}
surveys_middle = surveys[nrow(surveys)/2,]
```

4. Combine `nrow()` with the `-` notation above to reproduce the behavior of `head(surveys)`, keeping just the first through 6th rows of the surveys dataset.

```{r}
surveys[-(7:nrow(surveys)),]
```

# Factors

As seen in `str(surveys)`, many of the columns contained integers (`int`). Others, including `species_id`, `sex`, and `genus` are a special class called `factor`. Factors are categorical data stored as integers that are associated with labels instead of being stored as a character vector. 

Factors only contain a predefined set of values called `levels`. By default, R will sort levels in alphabetical order.

```{r}
sex = factor(c("male", "female", "female", "male"))
levels(sex) #returns the names of the levels in alphabetical order
nlevels(sex) #returns the number of levels
```

Sometimes, it is necessary to have the levels organized in a specific way that is not alphabetical (i.e. low, medium, high) for purposes of graphing or analysis. In these cases, the order can be manually changed:

```{r}
sex = factor(sex, levels = c("male", "female")) #manually orders the levels
```


# Converting factors 

You an also convert factors into character vectors (`as.character()`) or numerical vectors (`as.numeric()`). Converting factor levels to numbers that can be interpreted to do math can be tricky though. The function `as.numeric()` returns the index values of the factors in numerical order. To fix this we can convert the factors to characters first and then to numbers. Or we can use the `levels()` function which (1) returns the factor levels using `levels()`, (2) converts them to numbers using `as.numeric()`, and (3) access these numeric values using the underlying integers of the vector `year_fct` inside the square brackets

```{r}
as.character(sex) #converts a factor to a character vector
year_fct = factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct) #Wrong! And there is no warning...
as.numeric(as.character(year_fct)) #Works...
as.numeric(levels(year_fct))[year_fct] #The recommended way.
```

# Renaming factors

The `plot()` function can help you view the overall number of observations made corresponding to each factor level. 

```{r}
plot(as.factor(surveys$sex))
```

We see here that there are about 1700 individuals that do not have sex information recorded but they are also not labelled. We can manually change the `sex` value of all of these individuals by doing the following:

```{r}
sex <- factor(surveys$sex) #assigning the factors of the sex column to the variable 'sex'
head(sex)
levels(sex) #viewing the levels of the values in sex
levels(sex)[1] <- "undetermined" #changing the [1] level to "undetermined"
levels(sex) #viewing these changed values
head(sex)
```


### Challenge 3

Rename "F" and "M" to "female" and "male" respectively.

```{r}
levels(sex)[2] = "female"
levels(sex)[3] = "male"
levels(sex)
```

Now that we have renamed the factor level to “undetermined”, can you recreate the barplot such that “undetermined” is last (after “male”)?

```{r}
sex = factor(sex, levels = c("female", "male", "undetermined"))
plot(sex)
```

# Using `stringAsFactors = FALSE` 

When a data frame is imported using `read.csv()`, the columns containing characters are automatically converted into factors. To bypass this, we can use the argument `stringAsFactors = FALSE` when importing the data. 

```{r}
## Compare the difference between our data read as `factor` vs `character`.
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = TRUE)
str(surveys)
surveys <- read.csv("data/portal_data_joined.csv", stringsAsFactors = FALSE)
str(surveys)

## Convert the column "plot_type" into a factor
surveys$plot_type <- factor(surveys$plot_type)
```

### Challenge 4

1. We have seen how data frames are created when using `read.csv()`, but they can also be created by hand with the `data.frame()` function. There are a few mistakes in this hand-crafted `data.frame`. Can you spot and fix them? Don’t hesitate to experiment!

```{r eval = FALSE}
animal_data <- data.frame(
          animal = c(dog, cat, sea cucumber, sea urchin),
          feel = c("furry", "squishy", "spiny"),
          weight = c(45, 8 1.1, 0.8)
          )
```

**(1) The words in the array 'animal' should be set as characters, (2) There are only 3 values in the 'feel' array (there should be 4), (3) There is a missing comma between 8 and 1.1 in the 'weight' array.**

```{r}
animal_data <- data.frame(
          animal = c("dog", "cat", "sea cucumber", "sea urchin"),
          feel = c("furry", "squishy", "slimey", "spiny"),
          weight = c(45, 8, 1.1, 0.8)
          )
```

2.Can you predict the class for each of the columns in the following example? Check your guesses using str(country_climate):\
**Predictions: `country`, `climate`, `temperature`, and `northern_hemisphere` will be factors and `has_kangaroo` will be numeric.**

* Are they what you expected? Why? Why not?\
**Yes, any array containing a value in quotations will be a factor. The `has_kangaroo` column, the boolean values will be converted to 0s and 1s and be numeric.**
* What would have been different if we had added `stringsAsFactors = FALSE` when creating the data frame?\
**If we used `stringAsFActors = FALSE` they would become character vectors**
* What would you need to change to ensure that each column had the accurate data type?\
**You would need to remove the quotes from the two values in `northern_hemisphere` and `temperature`, and replace the 1 in `has_kangaroo` with a `TRUE`.**

```{r eval = FALSE}
country_climate <- data.frame(
       country = c("Canada", "Panama", "South Africa", "Australia"),
       climate = c("cold", "hot", "temperate", "hot/temperate"),
       temperature = c(10, 30, 18, "15"),
       northern_hemisphere = c(TRUE, TRUE, FALSE, "FALSE"),
       has_kangaroo = c(FALSE, FALSE, FALSE, 1)
       )
```

# Formatting dates

The best way to deal with date information is having each component of the date stored as separate variables. In this case, the date is stored in 3 different columns containing the year, month, and day as integers.

```{r}
str(surveys)
```

We can use the `ymd()` function from the `lubridate` package to create `Date` vectors. `Date` is a class of data recognized by R to be a date. It is esentially a character vector with the date in the format "YYY-MM-DD".

```{r results = 'hide'}
library("lubridate")
my_date <- ymd("2015-01-01")
str(my_date)
```

If we paste in the 3 separate values of the year, month, and day into the `ymd()` function separated by a "-" then we get the same results. Therefore, we can just use the `paste()` function with the 3 columns from the `surveys` dataset to format the dates.

```{r results = 'hide'}
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)

paste(surveys$year, surveys$month, surveys$day, sep = "-")
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

These values can now be added to the `surveys` dataset in a new column called `date`. We can inspect that this worked properly using `summary()` and `str()`.

```{r}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))

str(surveys)
summary(surveys$date)
```

Some of the date resulted in NAs. We can extract all the dates that are missing values and see what they have in common to figure out why they failed to parse. 

```{r}
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]
head(missing_dates)
```

**These dates failed to parse because they are non-exsistant dates. Either they must be changed so that they are existing dates or deleted entirely.**
