---
title: "Lab 4- Data wrangling and graphing COVID-19 reporting data"
author: "Briana Kubik"
date: "9/14/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Making a graph from the daily reports

```{r warning = FALSE}
library(tidyverse)
```



```{r}
report_03_11_2020 <-   read_csv(url("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_daily_reports/03-11-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State")
str(report_03_11_2020)
View(report_03_11_2020)
```

# Interactive data tables

The **`DT`** package can be used to make interactive tables.

```{r warning = FALSE}
library(DT)
datatable(report_03_11_2020)
```

### Exercises (Part 1)

Today’s exercises involve the application of what you learned in the Data Carpentries tutorial to a new data set. This step in applying your skills is often hard at first. Take your time and reread the tutorials, then try some options.

Use the DT package for showing the data tables where appropriate.

1. The format of the daily reports has changed over time. What are the differences between 3/13/2020, the Friday before spring break 2020 and yesterday 6 months later on 9/13/2020? Load the files into R and use `str()` and `View()` (or in the top right corner click on Environment and then the data frame.). Note the in the more recent dates they have changed the column headers so that it is no longer necessary to rename `Province_State` and `Country_Region`.

```{r}
report_03_13_2020 <-read_csv(url("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_daily_reports/03-13-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State")
str(report_03_13_2020)
#View(report_03_13_2020)

report_09_13_2020 <-  read_csv(url("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) 
str(report_09_13_2020)
#View(report_09_13_2020)
```

2. Many countries have multiple entries representing different regions (e.g. counties and states). To get the counts for a country we need to summarise the data. Use `group_by()` and `summarize()` to make a data frame (tibble) with total counts for each country.

```{r}
T_counts_3_13 = report_03_13_2020 %>% 
  group_by(Country_Region) %>% 
  summarize(Total_counts = n())

T_counts_9_13 = report_09_13_2020 %>% 
  group_by(Country_Region) %>% 
  summarize(Total_counts = n())
```

3. To get counts for confirmed cases per U.S state use `filter()` to select US and `group_by` and `summarize` as above.

```{r}
US_counts_3_13 = report_03_13_2020 %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_confirmed = sum(Confirmed))

US_counts_9_13 = report_09_13_2020 %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_confirmed = sum(Confirmed))
```

4. It will be difficult to make simple graphs with the data from all countries in the world or even the US states. Use `arrange_by()` and `slice()` to get just the top 10 countries for Deaths.

```{r}
T_death_3_13 = report_03_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_deaths = sum(Deaths)) %>% 
  arrange(desc(Total_deaths)) %>% 
  slice(1:10)

T_death_9_13 = report_09_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_deaths = sum(Deaths)) %>% 
  arrange(desc(Total_deaths)) %>% 
  slice(1:10)
```

# Introduction to R Graphics

Some more advanced and widely used graphing packages in R include **`grid`**, **`lattice`**, and **`ggplot2`**. We will be using **`ggplot2`** in this tutorial.

### Installation

**`tidayverse`** was installed and loaded previously so it is already in the working environment. Packages only need to be installed once and called to the workspace every time R is reopened.

### Tutorials and resources

Do the Data Carpentries "Data Visualization with ggplot2" tutorial in a separate file called Lab4a. 

### Exercises (Part 2)

1. Make plots using geom_point() of the 10 countries with the confirmed cases and deaths (two separate graphs) on 3/13/2020 and 9/13/2020.

```{r}
###Confirmed 3/2020###
T_3_13_confirmed = report_03_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_confirmed = sum(Confirmed)) %>% 
  arrange(desc(Total_confirmed)) %>% 
  slice(1:10)

ggplot(data = T_3_13_confirmed, mapping = aes(x = Country_Region, y = Total_confirmed)) +
  geom_point() +
  theme_bw() +
  labs(title = "Top 10 countries in confirmed COVID-19 cases (3/13/2020)",
       x = "Country",
       y = "Number of confirmed COVID-19 cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), text = element_text(size = 16),
       plot.title = element_text(size = 16, hjust = 0.5))

###Deaths 3/2020###
T_3_13_deaths = report_03_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_deaths = sum(Deaths)) %>% 
  arrange(desc(Total_deaths)) %>% 
  slice(1:10)

ggplot(data = T_3_13_deaths, mapping = aes(x = Country_Region, y = Total_deaths)) +
  geom_point() +
  theme_bw() +
  labs(title = "Top 10 countries in COVID-19 deaths (3/13/2020)",
       x = "Country",
       y = "Number of COVID-19 deaths") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), text = element_text(size = 16),
        plot.title = element_text(size = 16, hjust = 0.5))

###Confirmed 9/2020###
T_9_13_confirmed = report_09_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_confirmed = sum(Confirmed)) %>% 
  arrange(desc(Total_confirmed)) %>% 
  slice(1:10)

ggplot(data = T_9_13_confirmed, mapping = aes(x = Country_Region, y = Total_confirmed)) +
  geom_point() +
  theme_bw() +
  labs(title = "Top 10 countries in confirmed COVID-19 cases (9/13/2020)",
       x = "Country",
       y = "Number of confirmed COVID-19 cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), text = element_text(size = 16),
        plot.title = element_text(size = 16, hjust = 0.5))

###Deaths 9/2020###
T_9_13_deaths = report_09_13_2020 %>%
  group_by(Country_Region) %>% 
  summarize(Total_deaths = sum(Deaths)) %>% 
  arrange(desc(Total_deaths)) %>% 
  slice(1:10)

ggplot(data = T_9_13_deaths, mapping = aes(x = Country_Region, y = Total_deaths)) +
  geom_point() +
  theme_bw() +
  labs(title = "Top 10 countries in COVID-19 deaths (9/13/2020)",
       x = "Country",
       y = "Number of COVID-19 deaths") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10), text = element_text(size = 16),
        plot.title = element_text(size = 16, hjust = 0.5))
```


2. Make using geom_bar of the 10 states with the most confirmed cases and deaths (two separate graphs) on 3/12/2020 and 9/12/2020. 

```{r}
###Confirmed 3/2020###
US_3_13_confirmed = report_03_13_2020 %>%
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_confirmed = sum(Confirmed)) %>% 
  arrange(desc(US_state_confirmed)) %>% 
  slice(1:10)

ggplot(data = US_3_13_confirmed, mapping = aes(x = US_state_confirmed, y = Province_State)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  labs(title = "Top 10 US states in confirmed COVID-19 cases (3/13/2020)",
       x = "Number of confirmed COVID-19 cases",
       y = "State") +
  theme(axis.text.x = element_text(size = 12), text = element_text(size = 16),
        plot.title = element_text(size = 14, hjust = 0.5))

###Deaths 3/2020###
US_3_13_deaths = report_03_13_2020 %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_deaths = sum(Deaths)) %>% 
  arrange(desc(US_state_deaths)) %>% 
  slice(1:10)

ggplot(data = US_3_13_deaths, mapping = aes(x = US_state_deaths, y = Province_State)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  labs(title = "Top 10 US states in COVID-19 deaths (3/13/2020)",
       x = "Number of COVID-19 deaths",
       y = "State") +
  theme(axis.text.x = element_text(size = 12), text = element_text(size = 16),
        plot.title = element_text(size = 16, hjust = 0.5))

###Confirmed 9/2020###
US_9_13_confirmed = report_09_13_2020 %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_confirmed = sum(Confirmed)) %>% 
  arrange(desc(US_state_confirmed)) %>% 
  slice(1:10)

ggplot(data = US_9_13_confirmed, mapping = aes(x = US_state_confirmed, y = Province_State)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  labs(title = "Top 10 US states in confirmed COVID-19 cases (9/13/2020)",
       x = "Number of confirmed COVID-19 cases",
       y = "State") +
  theme(axis.text.x = element_text(size = 12), text = element_text(size = 16),
        plot.title = element_text(size = 14, hjust = 0.5))

###Deaths 3/2020###
US_9_13_deaths = report_09_13_2020 %>% 
  filter(Country_Region == "US") %>% 
  group_by(Province_State) %>% 
  summarize(US_state_deaths = sum(Deaths)) %>% 
  arrange(desc(US_state_deaths)) %>% 
  slice(1:10)

ggplot(data = US_9_13_deaths, mapping = aes(x = US_state_deaths, y = Province_State)) +
  geom_bar(stat = 'identity') +
  theme_bw() +
  labs(title = "Top 10 US states in COVID-19 deaths (9/13/2020)",
       x = "Number of COVID-19 deaths",
       y = "State") +
  theme(axis.text.x = element_text(size = 12), text = element_text(size = 16),
        plot.title = element_text(size = 16, hjust = 0.5))
```

